---
title: "Reproducible Research: Peer Assessment 1"
author: "Nate Brantley"
output: 
  html_document:
    keep_md: true
---

# Introduction
The purpose of this analysis is to analyze data from a personal activity monitoring device and determine if there are any differences in activity patterns between weekdays and weekends.

## Loading and preprocessing the data
The data I have utilized is contained in a file activity.csv. After extracting the .csv file from the .zip file, the data is loaded into a new data frame act_data.

```{r}
act_data <- read.table("activity.csv", header=TRUE, na.strings="NA", sep=",", quote="\"")
```
A review of the initial data set (using str function) is as follows:

```{r}
str(act_data)
```

The variables steps and interval are integers which is suitable. The interval variable is divided into 5-minute segments that re-set every hour. For example, interval 55 is 55 minutes after midnight; 100 is 1AM; 105 is 1:05AM; 2355 is 11:55PM.

The date variable is a factor which needs to be transformed into a date for proper analysis. I've also added a variable for day so that the data can be analyzed based on weekday vs. weekend.

```{r warning=FALSE, message=FALSE}
library(lubridate)
```
```{r}
act_data$date <- as_date(act_data$date)
act_data$day <- wday(act_data$date, label=TRUE)
```

The summary of act_data is now as follows:

```{r}
summary(act_data)
```
Perhaps most interesting is that there were 0 steps taken in more than half of all 5-minute intervals. Also noted is that of `r dim(act_data)[1]` observations, there were `r sum(is.na(act_data$steps))` missing values (`r sprintf("%1.1f%%", mean(is.na(act_data$steps))*100)` of the total).

A breakdown of missing values by day of week is as follows:

```{r warning=FALSE, message=FALSE}
library(dplyr)
```
```{r}
act_data %>% filter(is.na(steps)) %>%
            group_by(day) %>%
            summarize(no_na=sum(is.na(steps)))
```

## What is mean total number of steps taken per day?

To answer this question, I total the number of steps in the data set and then divide by the total number of days in the data set. **Note that this calculation excludes missing data.**

```{r}
round(sum(act_data$steps, na.rm=TRUE)/as.numeric(max(act_data$date)-min(act_data$date)+1), digits=0)
```

## What is the average daily activity pattern?

To examine the average daily activity pattern, I averaged the number of steps in each 5-minute interval across the study period with the results stored in act_data_by_interval. **Again, this calculation excludes missing data.**

```{r}
act_data_by_interval <- act_data %>% 
                        group_by(interval) %>%
                        summarize(avg_steps=mean(steps, na.rm=TRUE))
```

The daily activity pattern is plotted as follows:

```{r fig.height=4, fig.width=4, fig.align="center"}
plot(act_data_by_interval$interval, act_data_by_interval$avg_steps, type="l", main="No. of Steps by Interval\n Over Study Period", xlab="5-Minute Interval", ylab="Avg. No. of Steps", col="blue", lwd=2)
```

The interval with the highest number of average steps is determined as follows:

```{r}
filter(act_data_by_interval, avg_steps==max(avg_steps))
```

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
